### 2023-10-25 journal/introduction.tex
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
updated some related work

We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
to have step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations in the root are ordered, 
and this order is used to linearize the operations and compute their responses.
\here{Either here or in related work section, talk about previous usage of ordering tree and how ours differs from it}
Explicitly storing  operations in the tree nodes would be too costly.
Instead, we use a novel implicit representation of sets
of operations that allows us to quickly merge two sets from the children of a node,
and quickly access any  operation in a~set.
A preliminary version of this work appeared in~\cite{Nad22}.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
to have step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
Since a queue is also a bag, our queue is the first lock-free bag that we know of that has polylogarithmic step complexity.
Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations reach the root in batches and are linearized in the order they reach the root.
Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
or applying them one-by-one to the queue would be too costly.
Instead, we use a novel implicit representation of batches
of queue operations that allows us to quickly merge two batches from the children of a node,
and quickly access any  operation in a~batch.
A preliminary version of this work appeared in~\cite{Nad22}.
\here{Maybe say a little more about techniques used in the implementation, if space permits}

### 2023-10-25 journal/related.tex
updated some related work

Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} improved the MS-queue's performance using \emph{futures}.
Operations return future
objects instead of responses. Later, when an operation's response is needed, it
is evaluated using the future object.
This allows batches of enqueues or dequeues to be done at once on an MS-queue.
However, the implementation satisfies a weaker correctness condition than linearizability.
Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
to mix enqueues and dequeues.
\here{Is this worth saying:
They use some properties of the queue size before and after a batch, similar to a part of our work.}
In the worst case, where operations require their response right away,
batches have size 1, and both of these implementations behave like a standard MS-queue.

@@

\paragraph{Universal Constructions.}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
the universal construction of Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
(Thus, in terms of operations on reasonably-sized $O(\log p)$-bit words, their construction would take $\Omega(p\log p)$ steps per operation.)
Fatourou and Kallimanis \cite{FK14} used their own universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(p)$.
\here{Double check this.}

\paragraph{Restricted Queues.}
David gave the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but only for a single dequeuer. 
Our \ordering\ tree is similar to the tree structure they use to agree on a linearization ordering.
Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue  
that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.

\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?  May not be worth mentioning because then $p$ is a constant, so complexity in terms of $p$ doesn't make sense.}

\paragraph{Other Primitives.}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue 
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as %particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

\paragraph{Fetch\&Increment Objects.}
Ellen, Ramachandran and Woelfel \cite{ERW12} gave an implementation of 
fetch\&increment objects that uses a polylogarithmic number of steps per operation.
Like our queue, they also use a tree structure similar to the universal construction of 
\cite{DBLP:conf/stoc/AfekDT95} to keep track of the operations
that have been performed.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.
Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} gave an improved implementation of fetch\&increment with better step complexity.

\paragraph{Lower Bounds.}
As mentioned in the introduction, it follows from Attiya and Fouren's lower bound on bag data structures
\cite{DBLP:conf/opodis/AttiyaF17} that the
amortized step complexity of operations on a queue is $\Omega(\min(c,\log\log p))$, where $c$ is contention.
Subsequently, Jayanti, Tarjan and Boix-Adser\`{a} \cite{JTB19} proved a stronger $\Omega(\log p)$ lower bound on
the amortized step complexity of queue operations.

>>>>

Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} improved the MS-queue's performance using \emph{futures}.
Operations return future
objects instead of responses. Later, when an operation's response is needed, it
is evaluated using the future object.
This allows batches of enqueues or dequeues to be done at once on an MS-queue.
However, the implementation satisfies a weaker correctness condition than linearizability.
Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
to mix enqueues and dequeues.
%\here{Is this worth saying:
%They use some properties of the queue size before and after a batch, similar to a part of our work.}
In the worst case, where operations require their response right away,
batches have size 1, and both of these implementations behave like a standard MS-queue.

@@

\paragraph{Universal Constructions.}
\Eric{Rewrote this paragraph on Oct 25}
One can also build a queue using a universal construction \cite{10.1145/114005.102808}.
Afek, Dauber, and Touitou's universal construction~\cite{DBLP:conf/stoc/AfekDT95}
introduced the technique where a process must climb a binary tree while helping any other processes it sees along the way to reach the root,
and this technique is the basis of our queue implementation.
Jayanti \cite{DBLP:conf/podc/Jayanti98a} observed that
their construction can be modified to use $O(\log p)$ steps per operation, 
assuming that words can store $\Omega(p \log p)$ bits. 
However, if more reasonably-sized $O(\log p)$-bit words words are used, the construction would take $\Omega(p\log p)$ steps per operation.
There are two obstacles to making this construction more efficient:
it is expensive to manipulate lists of up to $p$ operations that must be propagated along a path
up to the root, and
when a batch of up to $p$ operations reaches the root, it is expensive to perform all of them on the
implemented data structure.
In our work, we devise a novel way of doing this implicitly for batches of operations on a queue.

Fatourou and Kallimanis \cite{FK14} used their own universal construction based on fetch\&add and LL/SC instructions
to implement a queue, but its step complexity is also $\Omega(p)$.
\here{Double check this.}

\paragraph{Restricted Queues.}
David gave the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.
Jayanti and Petrovic gave a wait-free polylogarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but only for a single dequeuer. 
Like our queue, theirs uses the cooperative tree-climbing
technique of Afek, Dauber and Touitou~\cite{DBLP:conf/stoc/AfekDT95}.
Concurrently with our work, which first appeared in \cite{Nad22}, Johnen, Khattabi and Milani \cite{JKM23} built on \cite{DBLP:conf/fsttcs/JayantiP05} to give a wait-free queue  
that achieves $O(\log p)$ steps for enqueue operations but fails to achieve polylogarithmic step complexity for dequeues: their dequeue operations take $O(k \log p)$ steps if there are $k$ dequeuers.

%\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?  May not be worth mentioning because then $p$ is a constant, so complexity in terms of $p$ doesn't make sense.}

\paragraph{Other Primitives.}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue 
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as %particular kinds of
double-word read-modify-write operations.
They use this as evidence to argue that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation undermines that argument.

\paragraph{Fetch\&Increment Objects.}
Ellen, Ramachandran and Woelfel \cite{ERW12} gave an implementation of 
fetch\&increment objects that uses a polylogarithmic number of steps per operation.
Like our queue, they also use a tree structure similar to the universal construction of 
\cite{DBLP:conf/stoc/AfekDT95} to keep track of the operations
that have been performed.
However, our construction requires more intricate
data structures to represent sets of operations, since a queue's state cannot be represented as succinctly
as the single-word state of a fetch\&increment object.
Ellen and Woelfel \cite{10.1007/978-3-642-41527-2_20} then gave an improved implementation of fetch\&increment with better step complexity.

\paragraph{Lower Bounds.}
As mentioned in the introduction, it follows from Attiya and Fouren's lower bound on bag data structures
\cite{DBLP:conf/opodis/AttiyaF17} that the
amortized step complexity of operations on a queue is $\Omega(\min(c,\log\log p))$, where $c$ is contention.
Subsequently, Jayanti, Tarjan and Boix-Adser\`{a} \cite{JTB19} proved a  $\Omega(\log c)$ lower bound on
the amortized step complexity of queue operations.
The combination of these lower bounds is shown schematically in Figure \ref{graph-lb}.
If we are interested in the complexity only as a function of $p$, the second result
gives us an $\Omega(\log p)$ lower bound, since $c$ can be as high as $p$.

\begin{figure}
\input{graph-lb.pdf_t}
\caption{Known lower bounds on amortized step complexity of queue operations when contention is~$c$ in a system of $p$ processes.\label{graph-lb}}
\end{figure}
